---
title: "R Spatial Tutorial for Conservation Biology"
subtitle: "Vector and raster analysis, species distribution mapping with sf, terra, ggplot2, and tmap"
author: "Conservation Biology Course"
date: today
---

## Introduction

This tutorial walks through a reproducible R workflow for analysing and
visualising spatial biodiversity data. We will:

1. Install and load the required packages.
2. Load and inspect vector and raster data.
3. Wrangle and summarise species occurrence records.
4. Create static maps with `ggplot2`.
5. Create interactive maps with `tmap`.
6. Perform a simple habitat-suitability overlay.

---

## 1 – Required packages

Install once (skip if already installed):

```{r}
#| label: install-packages
#| eval: false

install.packages(c(
  "sf",        # vector data (Simple Features)
  "terra",     # raster data
  "ggplot2",   # static mapping
  "tmap",      # thematic / interactive maps
  "dplyr",     # data wrangling
  "readr",     # fast CSV reading
  "viridis",   # colour palettes
  "ggspatial", # north arrow & scale bar for ggplot2
  "scales"     # axis label formatting (e.g. percent_format)
))
```

Load packages for this session:

```{r}
#| label: load-packages

library(sf)
library(terra)
library(ggplot2)
library(tmap)
library(dplyr)
library(readr)
library(viridis)
library(ggspatial)
library(scales)
```

---

## 2 – Loading spatial data

### 2.1 Study-area polygon (vector)

```{r}
#| label: load-study-area

study_area <- sf::st_read("data/study_area.gpkg", quiet = TRUE)
print(study_area)
```

Key attributes to check:

```{r}
#| label: check-crs

sf::st_crs(study_area)   # coordinate reference system
sf::st_bbox(study_area)  # bounding box
```

### 2.2 Species occurrence records (CSV)

```{r}
#| label: load-observations

obs_raw <- readr::read_csv("data/species_observations.csv", show_col_types = FALSE)
head(obs_raw)
```

Convert to an `sf` point object:

```{r}
#| label: csv-to-sf

obs <- sf::st_as_sf(
  obs_raw,
  coords = c("longitude", "latitude"),
  crs    = 4326          # WGS 84
)
print(obs)
```

### 2.3 Land-cover raster

```{r}
#| label: load-raster

lc <- terra::rast("data/land_cover.tif")
print(lc)
terra::plot(lc, main = "Land cover")
```

---

## 3 – Data wrangling

### 3.1 Reproject to a metric CRS

For distance- or area-based analyses, project all layers to the same metric CRS.
Here we use **UTM Zone 33N (EPSG:32633)**; adjust for your study region.

```{r}
#| label: reproject

target_crs <- 32633

study_area_utm <- sf::st_transform(study_area, crs = target_crs)
obs_utm        <- sf::st_transform(obs,        crs = target_crs)
lc_utm         <- terra::project(lc, paste0("EPSG:", target_crs),
                                 method = "near")  # categorical raster
```

### 3.2 Clip observations to the study area

```{r}
#| label: clip-observations

obs_inside <- obs_utm[sf::st_within(obs_utm, study_area_utm, sparse = FALSE), ]
cat("Observations inside study area:", nrow(obs_inside), "\n")
```

### 3.3 Summarise by species

```{r}
#| label: species-summary

species_count <- obs_inside |>
  dplyr::count(species, name = "n_observations") |>
  dplyr::arrange(dplyr::desc(n_observations))

print(species_count)
```

### 3.4 Extract land-cover values at observation points

```{r}
#| label: extract-land-cover

obs_lc <- obs_inside |>
  dplyr::mutate(
    land_cover_class = terra::extract(lc_utm, obs_inside)[, 2]
  )

table(obs_lc$land_cover_class, useNA = "ifany")
```

---

## 4 – Static maps with ggplot2

### 4.1 Basic observation map

```{r}
#| label: fig-obs-map
#| fig-cap: "Species observations inside the study area."

ggplot2::ggplot() +
  ggplot2::geom_sf(data = study_area_utm, fill = "grey90", colour = "grey40") +
  ggplot2::geom_sf(
    data   = obs_inside,
    ggplot2::aes(colour = species),
    size   = 2,
    alpha  = 0.8
  ) +
  viridis::scale_colour_viridis_d(option = "D", name = "Species") +
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +
  ggspatial::annotation_scale(location = "bl") +
  ggplot2::labs(
    title    = "Species observations in the study area",
    subtitle = "Conservation Biology Course",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_bw()
```

### 4.2 Observation density map

```{r}
#| label: fig-density-map
#| fig-cap: "Spatial density of all observations."

# Convert to a regular data frame for stat_density_2d
obs_coords <- obs_inside |>
  dplyr::mutate(
    x = sf::st_coordinates(obs_inside)[, 1],
    y = sf::st_coordinates(obs_inside)[, 2]
  ) |>
  sf::st_drop_geometry()

ggplot2::ggplot() +
  ggplot2::geom_sf(data = study_area_utm, fill = "grey95", colour = "grey40") +
  ggplot2::stat_density_2d_filled(
    data    = obs_coords,
    ggplot2::aes(x = x, y = y, fill = ggplot2::after_stat(level)),
    alpha   = 0.7,
    contour_var = "ndensity"
  ) +
  viridis::scale_fill_viridis_d(option = "B", name = "Density") +
  ggplot2::geom_sf(
    data  = obs_inside,
    size  = 1, shape = 3, colour = "white", alpha = 0.5
  ) +
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +
  ggspatial::annotation_scale(location = "bl") +
  ggplot2::labs(
    title = "Observation density",
    x = NULL, y = NULL
  ) +
  ggplot2::theme_bw()
```

### 4.3 Bar chart of observations per land-cover class

```{r}
#| label: fig-lc-barplot
#| fig-cap: "Number of observations per land-cover class."

obs_lc |>
  sf::st_drop_geometry() |>
  dplyr::filter(!is.na(land_cover_class)) |>
  dplyr::count(species, land_cover_class) |>
  ggplot2::ggplot(ggplot2::aes(x = land_cover_class, y = n, fill = species)) +
  ggplot2::geom_col(position = "dodge") +
  viridis::scale_fill_viridis_d(option = "C", name = "Species") +
  ggplot2::labs(
    title = "Observations per land-cover class",
    x     = "Land-cover class",
    y     = "Number of observations"
  ) +
  ggplot2::theme_bw()
```

---

## 5 – Interactive maps with tmap

`tmap` can render maps in interactive (leaflet) mode – useful for exploring data
before finalising figures.

```{r}
#| label: fig-tmap-interactive
#| fig-cap: "Interactive map of species observations."

tmap::tmap_mode("view")   # switch to interactive mode

tmap::tm_shape(study_area_utm) +
  tmap::tm_borders(col = "grey40", lwd = 1.5) +
  tmap::tm_fill(col = "grey90", alpha = 0.4) +
tmap::tm_shape(obs_inside) +
  tmap::tm_dots(
    col   = "species",
    size  = 0.05,
    alpha = 0.8,
    title = "Species",
    palette = "viridis"
  ) +
tmap::tm_layout(title = "Species observations")
```

Switch back to static mode for the remaining figures:

```{r}
#| label: tmap-static-mode

tmap::tmap_mode("plot")
```

---

## 6 – Habitat suitability overlay

A simple presence-only habitat-suitability index based on land-cover preference.

### 6.1 Calculate land-cover preference per species

```{r}
#| label: lc-preference

# Proportion of observations in each land-cover class per species
lc_pref <- obs_lc |>
  sf::st_drop_geometry() |>
  dplyr::filter(!is.na(land_cover_class)) |>
  dplyr::group_by(species) |>
  dplyr::count(land_cover_class) |>
  dplyr::mutate(proportion = n / sum(n)) |>
  dplyr::ungroup()

print(lc_pref)
```

### 6.2 Map preferred land-cover per species

```{r}
#| label: fig-lc-preference
#| fig-cap: "Land-cover preference per species (proportion of observations)."

lc_pref |>
  ggplot2::ggplot(ggplot2::aes(
    x    = land_cover_class,
    y    = proportion,
    fill = species
  )) +
  ggplot2::geom_col(position = "dodge") +
  viridis::scale_fill_viridis_d(option = "D", name = "Species") +
  ggplot2::scale_y_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    title = "Land-cover preference by species",
    x     = "Land-cover class",
    y     = "Proportion of observations"
  ) +
  ggplot2::theme_bw()
```

### 6.3 Reclassify raster to suitability score

As a simplified example, assign a suitability score (0–3) to each land-cover
class for one focal species:

```{r}
#| label: suitability-raster

# Example lookup: class -> suitability (0 = unsuitable, 3 = most suitable)
# Adjust class IDs to match your actual raster values
suitability_matrix <- matrix(
  c(1, 1,   # class 1 -> suitability 1 (e.g. agricultural land)
    2, 3,   # class 2 -> suitability 3 (e.g. forest)
    3, 2,   # class 3 -> suitability 2 (e.g. shrubland)
    4, 0),  # class 4 -> suitability 0 (e.g. urban)
  ncol = 2, byrow = TRUE
)

suitability <- terra::classify(lc_utm, suitability_matrix)
names(suitability) <- "suitability"
terra::plot(suitability, col = viridis::viridis(4), main = "Habitat suitability")
```

### 6.4 Suitability map with ggplot2

```{r}
#| label: fig-suitability-map
#| fig-cap: "Habitat suitability raster overlaid with species observations."

# Convert raster to data frame for ggplot2
suit_df <- terra::as.data.frame(suitability, xy = TRUE) |>
  dplyr::rename(score = suitability)

ggplot2::ggplot() +
  ggplot2::geom_raster(
    data = suit_df,
    ggplot2::aes(x = x, y = y, fill = score)
  ) +
  viridis::scale_fill_viridis_c(name = "Suitability", option = "D", na.value = "white") +
  ggplot2::geom_sf(data = study_area_utm, fill = NA, colour = "black", linewidth = 0.8) +
  ggplot2::geom_sf(
    data  = obs_inside,
    size  = 1.5, shape = 21,
    fill  = "white", colour = "red", stroke = 0.5
  ) +
  ggspatial::annotation_north_arrow(location = "tr", which_north = "true") +
  ggspatial::annotation_scale(location = "bl") +
  ggplot2::labs(
    title    = "Habitat suitability with species observations",
    subtitle = "White circles = observation points",
    x = NULL, y = NULL
  ) +
  ggplot2::coord_sf() +
  ggplot2::theme_bw()
```

---

## 7 – Saving outputs

```{r}
#| label: save-outputs
#| eval: false

# Export clipped observations
sf::st_write(obs_inside, "data/obs_inside_study_area.gpkg", delete_dsn = TRUE)

# Export suitability raster
terra::writeRaster(suitability, "data/habitat_suitability.tif", overwrite = TRUE)

# Save a ggplot figure
ggplot2::ggsave(
  filename = "figures/suitability_map.png",
  width  = 18, height = 14, units = "cm", dpi = 300
)
```

---

## Summary

In this tutorial you learned how to:

- Load vector and raster data into R with `sf` and `terra`.
- Reproject layers to a metric CRS.
- Clip, join, and summarise spatial data.
- Create publication-quality static maps with `ggplot2` and `ggspatial`.
- Build interactive maps with `tmap`.
- Perform a simple habitat-suitability reclassification.

Return to the [QGIS Tutorial](qgis_tutorial.qmd) if you want to see how the
same operations look in a graphical interface.
